---
title: "BNT162b2 Decieving Public or Saving Lives?"
author: 'Krish Doshi, Lauren Yee, Kian Sharifzadeh'
date: '`r Sys.Date()`'
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
- \usepackage{colortbl}
- \usepackage{float}
- \usepackage{graphicx}
- \usepackage{booktabs}
- \usepackage{hyperref}
- \usepackage{caption}
- \usepackage{subcaption}
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
#install.packages("janitor")
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
library(janitor)
library(fastR2)

data <- data.frame("Covid_Case" = c(8, 162, 170), "No_Covid_Case" = c(17403, 17349, 34752), "Total"= c(17411, 17511, 34922))

row.names(data) <- c("BNT162b2", "Placebo", "Total")

extended_df <- data.frame("Treatment"=c(rep("BNT162b2", 17411), rep("Placebo", 17511)), "Has_Covid" = c(rep("Yes", 8), rep("No", 17403), rep("Yes", 162), rep("No", 17349)))
```

# Abstract

Write your abstract here.

# Keywords

*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background

Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

## Visualization



```{r table}

table_test <- extended_df %>% 
  tabyl(Group, Has_Covid) %>%
    adorn_totals(where = c("row", "col")) %>%
    adorn_title()


```

```{r boxplot}
ggplot(data = extended_df, mapping=aes(x=Has_Covid, fill=Treatment)) + 
  geom_bar(position="fill")


```

```{r mosaic-plot}

mosaicplot(table_test, main = "Covid Cases", color = TRUE)

mosaicplot(~ Group + Has_Covid, data = extended_df, color = TRUE)

```

# Statistical Methods



## Model

## Make sure to prove why T is binomial in Appendix (see last slides for example slides)

```{r}
pfizer_estimator <- 8/17411
placebo_estimator <- 162 / 17511

pfizer_estimator
placebo_estimator

relative_risk <- pfizer_estimator / placebo_estimator
relative_risk

psi <- 1 - relative_risk
psi

```

From these calculations, we find that $\pi_{v} = \frac{8}{17411}$ and $\pi_{p} = \frac{162}{17511}$. In other words, $\pi_{v} \approx 0.00046$ and $\pi_{p} \approx 0.00925$.

Pfizer vaccine estimator: 8/17411

Placebo estimator: 162 / 17511

Total: 170 / 34922

Describe the statistical model used.

Let $T \sim Binom(170, \pi)$, where $\pi$ is equivalent to P(Vaccine|Infection). T denotes the number of COVID cases in the vaccine group out of all the sample cases. 170 represents the total number of COVID cases in the sample. Since $n_1$ is approximately equal to $n_2$, we can rewrite P(Vaccine|Infection) as $\pi = \frac{\pi_{v}}{\pi_{v} + \pi_{p}}$. Our parameter of interest is $\psi$, which is equal to $1 -\frac{\pi_v}{\pi_p} = \frac{\pi_p - \pi_v}{\pi_p}$. Using our value of $\pi$, we can rewrite $\psi$ as $\frac{1-2\pi}{1-\pi}$.

Proof: 

$\frac{1-2\pi}{1-\pi} = \frac{1 - 2(\frac{\pi_{v}}{\pi_{v} + \pi_{p}})}{1- \frac{\pi_{v}}{\pi_{v} + \pi_{p}}}$

$\frac{\pi_v + \pi_p - 2\pi_v}{\pi_v + \pi_p - \pi_v}$

$\frac{\pi_p - \pi_v}{\pi_p} = \psi$

Let $c$ be the number of COVID infections from the vaccine group. We observe $t = 8$. We want to test: $H_0: \psi = 0.3$, $H_1: \psi \ne 0.3$.



## Likelihood Inference

Detail the likelihood approach.

First, we need to find our likelihood function $L^{\ast}(\psi)$. Using our parameterization $\frac{1-2\pi}{1-\pi} = \psi$:

$1-2\pi = \psi(1-\pi)$

$1-2\pi = \psi-\psi\pi$

$\psi\pi-2\pi = \psi- 1$

$\pi(\psi-2) = \psi-1$

$\pi = \frac{\psi- 1}{\psi-2}$

We know that $L^{\ast}(\pi)  = \binom{n}{x} (\pi)^{x} \times (1-\pi)^{n-x}$. In this case, $n$ represents the number of people included in the sample and $x$ is the number of people who got infected with COVID. Using $\pi$ in terms of $\psi$, we get: $L^{\ast}(\psi) = \binom{n}{x} (\frac{\psi- 1}{\psi-2})^{x} \times (1-\frac{\psi- 1}{\psi-2})^{n-x}$

Next, we solve for $\hat{\psi}^{mle}$. In order to do this, we need to find the log-likelihood function of $L^{\ast}(\psi)$: 

$\ell(\psi) = \ln{\left[\binom{n}{x} (\frac{\psi- 1}{\psi-2})^{x}\right]} + \ln\left[{(1-\frac{\psi- 1}{\psi-2})^{n-x}}\right]$

$\ell(\psi) = \ln{\binom{n}{x}} + x\ln{(\psi- 1)} -x\ln{(\psi-2)} + (n-x)\ln{(\frac{\psi- 2 - (\psi- 1)}{\psi-2})}$

$\ell(\psi) = \ln{\binom{n}{x}} + x\ln{(\psi- 1)} -x\ln{(\psi-2)} + (n-x)\ln{(\frac{-1}{\psi-2})}$

Now, we take the first derivative with respect to $\psi$:

$\frac{d}{d \psi}\ell(\psi) = 0 + \frac{x}{\psi - 1} - \frac{x}{\psi - 2} - \frac{n-x}{\psi-2} = \frac{x}{\psi - 1} - \frac{x + (n-x)}{\psi - 2}$

$\frac{d}{d \psi}\ell(\psi) = \frac{x}{\psi - 1} - \frac{n}{\psi - 2}$

Setting $\frac{d}{d \psi}\ell(\psi) = 0$: $\frac{x}{\psi - 1} - \frac{n}{\psi - 2} = 0$

$\frac{x}{\psi - 1} = \frac{n}{\psi - 2}$

$x\psi - 2x = n\psi - n$

$n- 2x = \psi(n - x)$

Therefore, $\boxed{\hat{\psi}^{mle} = \frac{n - 2x}{n-x}}$, which can also be rewritten as $\frac{1 - 2\hat{\pi}^{mle}}{1 - \hat{\pi}^{mle}}$.

To see if $\hat{\psi}^{mle}$, we do the 2nd derivative test:

$\frac{d^2}{d \psi^2}\ell(\psi) = -\frac{x}{(\psi - 1)^2} + \frac{n}{(\psi - 2)^2}$

Since $\frac{n}{\psi - 2}$ will always be less than $\frac{x}{\psi - 1}$, $\hat{\psi}^{mle}$ is a \fbox{Strict Global Maximum}.

## Bayesian Inference

Detail the Bayesian approach.

# Results

## Frequentist Way

The equation for the large sample confidence interval of $\psi$ is $\hat{\psi}^{mle} \pm z_{\alpha/2} \hat{SE}$, where $\hat{SE} = \frac{1}{\sqrt{I(\hat{\psi}^{mle})}}$

Using Theorem 12.1 under regularity conditions, we know that $I(\hat{\psi}^{mle}) = -E\left[\frac{d^2}{d\psi^2}\ell(\psi)\right]$

$-E\left[-\frac{t}{(\psi - 1)^2} + \frac{n}{(\psi - 2)^2} \right] = \frac{t}{(\psi - 1)^2} - \frac{n}{(\psi - 2)^2} $

$\ell(\psi) = \ln{\binom{n}{x}} + x\ln{(\psi- 1)} -x\ln{(\psi-2)} + (n-x)\ln{(\frac{-1}{\psi-2})}$

```{r}


loglik.psi<-function(pi,x, n){ #first argument must be parameter
ifelse((pi < 0 | pi > 1), NA,
x*log((1-2*pi) / (1-pi))+(n-x)*log((1-((1-2*pi) / (1-pi))) ) )
}
#annoying: loglik argument is lowercase "l"
ml.psi <- maxLik2(loglik=loglik.psi,
start = 0.3, x = 8,n =170)
plot(ml.psi) +
labs(title = "Second order approximation for Psi Reparametrization",
subtitle = "n = 170, x = 8)",
x = expression(pi))



```


Present your findings.

# Discussion / Conclusion

Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from <https://www.example.com>.

# Appendix

## Code

```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs

If applicable, include detailed mathematical derivations or additional theoretical explanations.
