---
title: "BNT162b2 Decieving Public or Saving Lives?"
author: 'Krish Doshi, Lauren Yee, Kian Sharifzadeh'
date: '`r Sys.Date()`'
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
- \usepackage{colortbl}
- \usepackage{float}
- \usepackage{graphicx}
- \usepackage{booktabs}
- \usepackage{hyperref}
- \usepackage{caption}
- \usepackage{subcaption}
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
#install.packages("janitor")
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
library(janitor)

data <- data.frame("Covid_Case" = c(8, 162, 170), "No_Covid_Case" = c(17403, 17349, 34752), "Total"= c(17411, 17511, 34922))

row.names(data) <- c("BNT162b2", "Placebo", "Total")

extended_df <- data.frame("Treatment"=c(rep("BNT162b2", 17411), rep("Placebo", 17511)), "Has_Covid" = c(rep("Yes", 8), rep("No", 17403), rep("Yes", 162), rep("No", 17349)))
```

# Abstract

Write your abstract here.

# Keywords

*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background

Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

## Visualization



```{r table}

table_test <- extended_df %>% 
  tabyl(Group, Has_Covid) %>%
    adorn_totals(where = c("row", "col")) %>%
    adorn_title()


```

```{r boxplot}
ggplot(data = extended_df, mapping=aes(x=Has_Covid, fill=Treatment)) + 
  geom_bar(position="fill")


```

```{r mosaic-plot}

mosaicplot(table_test, main = "Covid Cases", color = TRUE)

mosaicplot(~ Group + Has_Covid, data = extended_df, color = TRUE)

```

# Statistical Methods



## Model

## Make sure to prove why T is binomial in Appendix (see last slides for example slides)

```{r}
pfizer_estimator <- 8/17411
placebo_estimator <- 162 / 17511

pfizer_estimator
placebo_estimator

relative_risk <- pfizer_estimator / placebo_estimator
relative_risk

psi <- 1 - relative_risk
psi

```

From these calculations, we find that $\pi_{v} = \frac{8}{17411}$ and $\pi_{p} = \frac{162}{17511}$. In other words, $\pi_{v} \approx 0.00046$ and $\pi_{p} \approx 0.00925$.

Pfizer vaccine estimator: 8/17411

Placebo estimator: 162 / 17511

Total: 170 / 34922

Describe the statistical model used.

Let $T \sim Binom(170, \pi)$, where $\pi$ is equivalent to P(Vaccine|Infection). T denotes the number of COVID cases in the vaccine group out of all the sample cases. 170 represents the total number of COVID cases in the sample. Since $n_1$ is approximately equal to $n_2$, we can rewrite P(Vaccine|Infection) as $\pi = \frac{\pi_{v}}{\pi_{v} + \pi_{p}}$. Our parameter of interest is $\psi$, which is equal to $1 -\frac{\pi_v}{\pi_p} = \frac{\pi_p - \pi_v}{\pi_p}$. Using our value of $\pi$, we can rewrite $\psi$ as $\frac{1-2\pi}{1-\pi}$.

Proof: 

$\frac{1-2\pi}{1-\pi} = \frac{1 - 2(\frac{\pi_{v}}{\pi_{v} + \pi_{p}})}{1- \frac{\pi_{v}}{\pi_{v} + \pi_{p}}}$

$\frac{\pi_v + \pi_p - 2\pi_v}{\pi_v + \pi_p - \pi_v}$

$\frac{\pi_p - \pi_v}{\pi_p} = \psi$

Let $c$ be the number of COVID infections from the vaccine group. We observe $t = 8$. We want to test: $H_0: \psi = 0.3$, $H_1: \psi \ne 0.3$.



## Likelihood Inference

Detail the likelihood approach.



## Bayesian Inference

Detail the Bayesian approach.

# Results

Present your findings.

# Discussion / Conclusion

Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from <https://www.example.com>.

# Appendix

## Code

```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs

If applicable, include detailed mathematical derivations or additional theoretical explanations.
