---
title: "BNT162b2 Decieving Public or Saving Lives?"
author: 'Krish Doshi, Lauren Yee, Kian Sharifzadeh'
date: '`r Sys.Date()`'
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
```

# Abstract

Write your abstract here.

# Keywords

*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background

Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

# Statistical Methods

## Model

```{r}
pfizer_estimator <- 8/17411
placebo_estimator <- 162 / 17511

pfizer_estimator
placebo_estimator

relative_risk <- pfizer_estimator / placebo_estimator
relative_risk

psi <- 1 - relative_risk
psi

```

From these calculations, we find that $\pi_{pf} = \frac{8}{17411}$ and $\pi_{pl} = \frac{162}{17511}$. In other words, $\pi_{pf} \approx 0.00046$ and $\pi_{pl} \approx 0.00925$.

Pfizer vaccine estimator: 8/17411

Placebo estimator: 162 / 17511

Total: 170 / 34922

Describe the statistical model used.

Let $T \sim Binom(170, \pi)$, where $\pi$ is equivalent to P(Vaccine|Infection). Since $n_1$ is approximately equal to $n_2$, we can rewrite P(Vaccine|Infection) as $\pi = \frac{\pi_{pf}}{\pi_{pf} + \pi_{pl}}$





## Likelihood Inference

Detail the likelihood approach.



## Bayesian Inference

Detail the Bayesian approach.

# Results

Present your findings.

# Discussion / Conclusion

Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from <https://www.example.com>.

# Appendix

## Code

```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs

If applicable, include detailed mathematical derivations or additional theoretical explanations.
