---
title: "BNT162b2 Deceiving Public or Saving Lives?"
author: 'Krish Doshi, Lauren Yee, Kian Sharifzadeh'
date: '`r Sys.Date()`'
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
- \usepackage{colortbl}
- \usepackage{float}
- \usepackage{graphicx}
- \usepackage{booktabs}
- \usepackage{hyperref}
- \usepackage{caption}
- \usepackage{subcaption}
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
#install.packages("janitor")
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
library(janitor)
library(fastR2)

data <- data.frame("Covid_Case" = c(8, 162, 170), "No_Covid_Case" = c(17403, 17349, 34752), "Total"= c(17411, 17511, 34922))

row.names(data) <- c("BNT162b2", "Placebo", "Total")

extended_df <- data.frame("Treatment"=c(rep("BNT162b2", 17411), rep("Placebo", 17511)), "Has_Covid" = c(rep("Yes", 8), rep("No", 17403), rep("Yes", 162), rep("No", 17349)))
```

# Abstract

Write your abstract here.

# Keywords

*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background

Provide an introduction with background information.

On March 11, 2020, Coronavirus disease 2019 (Covid-19) was declared a pandemic by the World Health Organization. The disease affected tens of millions of people worldwide and urged the production of safe and effective vaccines to prevent additional consequences. Shortly after the declaration, major pharmaceutical companies raced to develop vaccines against the virus, including the Pfizer-BioNTech vaccine, BNT162b2 examined in this report. Beyond its unprecedented use of mRNA and pace of development in about 10 months compared to the typical vaccine that takes 10 to 15 years, the efficacy rate easily passed the FDA standard. Specifically, the original article states that the "95.0% credible interval for vaccine efficacy and the probability of vaccine efficacy greater than 30% were calculated with the use of a Bayesian beta-binomial model". To explore the BNT162b2 vaccine findings further, we conduct analyses with Frequentist and Bayesian methods for inference.

https://www.nejm.org/doi/full/10.1056/nejmoa2034577
https://www.cdc.gov/museum/timeline/covid19.html
https://pmc.ncbi.nlm.nih.gov/articles/PMC7937046/#:~:text=Abstract,the%20governments%20across%20the%20world.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

## Visualization

```{r table}
table_test <- extended_df %>% 
tabyl(Treatment, Has_Covid) %>%
adorn_totals(where = c("row", "col")) %>%
adorn_title()
```

``` {r barplot}
ggplot(data = extended_df, mapping=aes(x=Has_Covid, fill=Treatment)) +
  geom_bar(position="fill") + 
  scale_fill_manual(values = c("lightblue", "#F5F5F5")) + 
  theme_minimal()

```

# Statistical Methods

## Model

## Make sure to prove why T is binomial in Appendix (see last slides for example slides)

```{r}
pfizer_estimator <- 8/17411
placebo_estimator <- 162 / 17511

pfizer_estimator
placebo_estimator

relative_risk <- pfizer_estimator / placebo_estimator
relative_risk

psi <- 1 - relative_risk
psi

```

From these calculations, we find that $\pi_{v} = \frac{8}{17411}$ and $\pi_{p} = \frac{162}{17511}$. In other words, $\pi_{v} \approx 0.00046$ and $\pi_{p} \approx 0.00925$.

Pfizer vaccine estimator: 8/17411

Placebo estimator: 162 / 17511

Total: 170 / 34922

With the statistical model, we let $T \sim Binom(170, \pi)$, where $\pi$ is equivalent to P(Vaccine|Infection). T denotes the number of COVID cases in the vaccine group out of all the sample cases. 170 represents the total number of COVID cases in the sample. Since $n_1$ is approximately equal to $n_2$, we can rewrite P(Vaccine|Infection) as $\pi = \frac{\pi_{v}}{\pi_{v} + \pi_{p}}$. Our parameter of interest is $\psi$, which is equal to $1 -\frac{\pi_v}{\pi_p} = \frac{\pi_p - \pi_v}{\pi_p}$. Using our value of $\pi$, we can rewrite $\psi$ as $\frac{1-2\pi}{1-\pi}$.

Proof: 

$\frac{1-2\pi}{1-\pi} = \frac{1 - 2(\frac{\pi_{v}}{\pi_{v} + \pi_{p}})}{1- \frac{\pi_{v}}{\pi_{v} + \pi_{p}}}$

$\frac{\pi_v + \pi_p - 2\pi_v}{\pi_v + \pi_p - \pi_v}$

$\frac{\pi_p - \pi_v}{\pi_p} = \psi$

Let $t$ be the number of COVID infections from the vaccine group. We observe $t = 8$. We want to test: $H_0: \psi = 0.3$, $H_1: \psi \ne 0.3$.

## Likelihood Inference

First, we find our likelihood function $L^{\ast}(\psi)$. Using our parameterization $\frac{1-2\pi}{1-\pi} = \psi$:

$1-2\pi = \psi(1-\pi)$

$1-2\pi = \psi-\psi\pi$

$1-\psi = \pi(2 -\psi)$

$\pi = \frac{1 -\psi}{2 -\psi}$

We know that $L^{\ast}(\pi)  = \binom{n}{t} (\pi)^{t} \times (1-\pi)^{n-t}$. In this case, $n$ represents the number of people included in the sample and $t$ is the number of people who got infected with COVID. Using $\pi$ in terms of $\psi$, we get: $L^{\ast}(\psi) = \binom{n}{t} (\frac{1 -\psi}{2 -\psi})^{t} \times (1-\frac{1 -\psi}{2 -\psi})^{n-t}$

Next, we solve for $\hat{\psi}^{mle}$. In order to do this, we need to find the log-likelihood function of $L^{\ast}(\psi)$: 

$\ell(\psi) = \ln{\left[\binom{n}{t} (\frac{1 -\psi}{2 -\psi})^{t}\right]} + \ln\left[{(1-\frac{1 -\psi}{2 -\psi})^{n-t}}\right]$

$\ell(\psi) = \ln{\binom{n}{t}} + t\ln{(1-\psi)} -t\ln{(2 - \psi)} + (n-t)\ln{(\frac{2-\psi - (1-\psi)}{2-\psi})}$

$\ell(\psi) = \ln{\binom{n}{t}} + t\ln{(1-\psi)} -t\ln{(2 - \psi)}  + (n-t)\ln{1} - (n-t)\ln{(2-\psi)}$

Now, we take the first derivative with respect to $\psi$:

$\frac{d}{d \psi}\ell(\psi) = 0 - \frac{t}{1-\psi} + \frac{t}{2-\psi} + \frac{n-t}{2-\psi} = -\frac{t}{1-\psi} + \frac{t + (n-t)}{2-\psi}$

$\frac{d}{d \psi}\ell(\psi) = -\frac{t}{1-\psi} + \frac{n}{2-\psi}$

Setting $\frac{d}{d \psi}\ell(\psi) = 0$: $-\frac{t}{1-\psi} + \frac{n}{2-\psi} = 0$

$\frac{t}{1-\psi} = \frac{n}{2-\psi}$

$2t - t\psi = n - n\psi$

$n- 2t = \psi(n - t)$

Therefore, $\boxed{\hat{\psi}^{mle} = \frac{n - 2t}{n-t}}$, which can also be rewritten as $\frac{1 - 2\hat{\pi}^{mle}}{1 - \hat{\pi}^{mle}}$.

To see if $\hat{\psi}^{mle}$, we perform the 2nd derivative test:

$\frac{d^2}{d \psi^2}\ell(\psi) = \frac{t}{(1-\psi)^2} -\frac{n}{(2-\psi)^2}$

Depending on the values of $n$ and $t$, our estimate will not always be considered a global maximum. In our case, $\frac{d^2}{d \psi^2}\ell(\psi)$ is negative, which means $\hat{\psi}^{mle}$ is a \fbox{Maximum}.

## Bayesian Inference

In the paper, the authors mention that they used a beta binomial model with the transformation $$\pi = \frac{\pi_v}{\pi_v + \pi_p}$$ where $\pi_v$ and $\pi_p$ are the probabilities of an infection in the vaccine and placebo groups respectively. Here, the binomial $\pi$ was modeled with a beta prior, Beta(0.700102, 1), and the parameter of interest was $\psi = \frac{1 - 2\pi}{1 - \pi}$. It was estimated to be 95% with the 95% Bayesian credible interval reported as [90.3%, 97.6%] and the posterior probability that vaccine efficacy exceeded 30% was larger than 0.999.

While the original paper used a minimally informative Beta(0.700102, 1) prior, we further investigate the robustness of this conclusion by utilizing several alternative prior beliefs based on two apriori about $\psi$.

$$P(\psi \geq 0.3) = P\left(\frac{1 - 2\pi}{1 - \pi} \geq 0.3\right)$$
$$= P(1 - 2\pi \geq 0.3(1 - \pi))$$
$$= P(1 - 2\pi \geq 0.3 - 0.3\pi)$$
$$= P(-1.7\pi \geq -0.7)$$
$$= P\left(\pi \leq \frac{7}{17}\right)$$
1 - Most Pessimistic

- $P(\psi \geq 0.3) = 0.025$
- $P(\psi \geq 0) = 0.35$

2 - Pessimestic

- $P(\psi \geq 0.3) = 0.05$
- $P(\psi \geq 0) = 0.5$

3 - Neutral

- $P(\psi \geq 0.3) = 0.1$
- $P(\psi \geq 0) = 0.65$

4 - Optimistic

- $P(\psi \geq 0.3) = 0.2$
- $P(\psi \geq 0) = 0.8$

5 - More Optimistic

- $P(\psi \geq 0.3) = 0.4$
- $P(\psi \geq 0) = 0.95$

```{r}
library(LearnBayes)

beta.select(quantile1 = list(p = 0.025, x = 7/17),
            quantile2 = list(p = 0.35, x = 0.5))

beta.select(quantile1 = list(p = 0.05, x = 7/17),
            quantile2 = list(p = 0.5, x = 0.5))

beta.select(quantile1 = list(p = 0.1, x = 7/17),
            quantile2 = list(p = 0.65, x = 0.5))

beta.select(quantile1 = list(p = 0.2, x = 7/17),
            quantile2 = list(p = 0.8, x = 0.5))

beta.select(quantile1 = list(p = 0.4, x = 7/17),
            quantile2 = list(p = 0.95, x = 0.5))
```
We choose these apriori beliefs to analyze the sensitivity of the prior. More specifically, these five priors represent varying degrees of pessimism and optimism regarding whether the vaccine exceeds the FDA efficacy threshold of 30% and whether it provides any beneficial effect.

Next, we find the posterior distributions and Bayesian p-values for the five priors with data model $f(T|\pi) \sim Binom(170, \pi)$. 

For all five prior, the hypotheses are:

$H_0: \psi = 0.3$ vs $H_1: \psi > 0.3$ where $\psi$ denotes the vaccine efficacy

Since $\psi \geq 0.3 \Leftrightarrow \pi \leq \frac{7}{17}$, we get that $P(\psi  \leq 0.3|t) = P\left(\pi \geq \frac{7}{17}|t\right)$.

To calculate the Bayesian probabilities: $P(\pi \geq \frac{7}{17}|t) = 1 - P(\pi \leq \frac{7}{17}|t)$ where $t = 8$.

1 - Most Pessimistic

$$g(\pi) = Beta(41.05, 37.63) \quad prior$$
$$h(\pi|t) = Beta(41.05 + t, 170 - t + 37.63) \quad posterior$$

p-val: $1 - P(\pi \leq \frac{7}{17}|t = 8)$ where $\pi|t \sim Beta(49.05, 199.63) \approx 0$

2 - Pessimistic

$$g(\pi) = Beta(43.03, 43.03) \quad prior$$
$$h(\pi|t) = Beta(43.03 + t, 170 - t + 43.03) \quad posterior$$

p-val: $1 - P(\pi \leq \frac{7}{17}|t = 8)$ where $\pi|t \sim Beta(51.03, 205.03) \approx 0$

3 - Neutral

$$g(\pi) = Beta(42.4, 46.0) \quad prior$$
$$h(\pi|t) = Beta(42.4 + t, 170 - t + 46.0) \quad posterior$$

p-val: $1 - P(\pi \leq \frac{7}{17}|t = 8)$ where $\pi|t \sim Beta(50.4, 208) \approx 0$

4 - Optimistic

$$g(\pi) = Beta(41.23, 49.20) \quad prior$$
$$h(\pi|t) = Beta(41.23 + t, 170 - t + 49.20) \quad posterior$$

p-val: $1 - P(\pi \leq \frac{7}{17}|t = 8)$ where $\pi|t \sim Beta(49.23, 211.2) \approx 0$

5 - More Optimistic

$$g(\pi) = Beta(48.96, 66.55) \quad prior$$
$$h(\pi|t) = Beta(48.96 + t, 170 - t + 66.55) \quad posterior$$

p-val: $1 - P(\pi \leq \frac{7}{17}|t = 8)$ where $\pi|t \sim Beta(56.96, 228.55) \approx 0$

```{r bay_pvals}
# Calculating Bayesian P-values

x = 7/17
n = 170
t = 8

most_pess <- pbeta(q = x, shape1 = 41.05 + t, shape2 = n - t + 37.63, lower.tail = F)

pess <- pbeta(q = x, shape1 = 43.03 + t, shape2 = n - t + 43.03, lower.tail = F)

neutral <- pbeta(q = x, shape1 = 42.4 + t, shape2 = n - t + 46, lower.tail = F)

opt <- pbeta(q = x, shape1 = 41.23 + t, shape2 = n - t + 49.2, lower.tail = F)

most_opt <- pbeta(q = x, shape1 = 48.96 + t, shape2 = n - t + 66.55, lower.tail = F)

most_pess
pess
neutral
opt
most_opt
```

The resulting Bayesian P-values are essentially 0. This indicates extremely strong posterior evidence that the vaccine efficacy exceeds the FDA threshold of 30%. The conclusion is therefore robust to the choice of prior, whether it is overly pessimistic or optimistic.

# Results

## Frequentist Way

### Confidence Interval

The equation for the large sample confidence interval of $\psi$ is $\hat{\psi}^{mle} \pm z_{\alpha/2} \hat{SE}$, where $\hat{SE} = \frac{1}{\sqrt{I(\hat{\psi}^{mle})}}$

Using Theorem 12.1 under regularity conditions, we know that $I(\hat{\psi}^{mle}) = E\left[-\frac{d^2}{d\psi^2}\ell(\psi)\right]$

$\frac{d^2}{d \psi^2}\ell(\psi) = -\frac{t}{(1-\psi)^2} + \frac{n}{(2-\psi)^2}$

$E\left[-(-\frac{T}{(\psi - 1)^2} + \frac{n}{(\psi - 2)^2} )\right] = \frac{E\left[T\right]}{(1-\psi)^2} - \frac{n}{(2-\psi)^2}$

$E[T] = n\pi = n\frac{1-\psi}{2-\psi}$

$I(\hat{\psi}^{mle}) = n\frac{1-\psi}{2-\psi} \cdot \frac{1}{{(1-\psi)^2}} - \frac{n}{(2-\psi)^2} = \frac{n}{(2-\psi)(1-\psi)}   - \frac{n}{(2-\psi)^2}$

$\frac{n(2-\psi)}{(2-\psi)^2(1-\psi)} - \frac{n(1-\psi)}{(2-\psi)^2(1-\psi)} = \frac{n}{(2-\psi)^2(1-\psi)} = I(\hat{\psi}^{mle})$

Since we got $I(\hat{\psi}^{mle})$, our standard error is $\frac{1}{\sqrt{\frac{n}{(2-\psi)^2(1-\psi)}}} = \sqrt{\frac{(2-\psi)^2(1-\psi)}{n}}$.

Using all the information given, we can solve the 95\% confidence interval with $t = 8$ and $n = 170$: 

```{r}
n <- 170
t <- 8
psi_mle <- (n - 2*t) / (n - t)
psi_se <- sqrt(((1-psi_mle)*(2-psi_mle)^2) / n)

lower <- psi_mle - qnorm(0.975)*psi_se
upper <- psi_mle + qnorm(0.975)*psi_se

lower
upper
psi_se 
```

$\ell(\psi) = \ln{\binom{n}{t}} + t\ln{(1-\psi)} -t\ln{(2 - \psi)}  + (n-t)\ln{(\frac{1}{2-\psi})}$
```{r}
loglik.psi<-function(psi,x, n){
  if(psi >= 1) NA
  else{  
    log(choose(n, x)) + x*log(1-psi) - x*log(2-psi) +(n-x)*log(1 / (2-psi))
  }
}

ml.psi <- maxLik2(loglik=loglik.psi,
start = 0.3, x = 8,n = 170)

plot(ml.psi) +
labs(title = "Second order approximation for Psi (Parametrization for Pi)",
subtitle = "n = 170, t = 8",
x = expression(psi))



loglik.pi<-function(pi,x, n){
  if(pi < 0 || pi > 1) NA
  else{  
    log(choose(n, x)) + x*log(pi) +(n-x)*log(1 - pi)
  }
}

ml.pi <- maxLik2(loglik=loglik.pi,
start = 0.3, x = 8,n = 170)

plot(ml.pi) +
labs(title = "Second order approximation for Pi",
subtitle = "n = 170, t = 8",
x = expression(pi))
```

```{r}
set.seed(414)

pi_samples <- rbinom(n = 100000, size = 170, prob =  8/170)
pi_mle <- pi_samples / 170
psi_samples <- (1 - 2*pi_mle) / (1-pi_mle)
sd(psi_samples)

```

### Likelihood Ratio Test Statistic

The likelihood ratio statistic is calculated by $W = 2\ln\Lambda$, which is equivalent to $2\ln(\frac{L(\hat{\psi}^{mle})}{L(\hat{\psi}^{null})}) = 2\left[\ln{L(\hat{\psi}^{mle})} - \ln{L(\hat{\psi}^{null})}\right]$. Given that $\hat{\psi}^{mle} \approx 0.9506$ and $H_0: \psi = 0.3$, we are solving for $2\left[\ell{(0.9506)} - \ell{(0.3)}\right]$. Using that $t = 8$ and $n = 160$, we can calculation $\ell(0.9506)$ and $\ell(0.3)$:

$\ell(0.9506) = \ln{\binom{160}{8}} + 8\ln{(1-0.9506)} -8\ln{(2 - 0.9506)}  + (162)\ln{(\frac{1}{2-0.9506})} \approx -1.945$

$\ell(0.3) = \ln{\binom{160}{8}} + 8\ln{(0.7)} -8\ln{(1.3)}  + (162)\ln{(\frac{1}{1.3})} \approx - 62.746$

Then, $W = 2(-1.945 + 62.746) \approx \boxed{121.601}$


Likelihood Ratio Test Statistic
```{r}
log_lik_func <- function(psi, n , t){
  log(choose(n, t)) + t*log(1-psi) - t*log(2-psi) -(n-t)*log((2-psi))
}

log_mle <- log_lik_func(psi_mle, 170, 8)
log_null <- log_lik_func(0.3, 170, 8)

log_mle
log_null

W_stat <- 2*(log_mle - log_null)
W_stat

```

```{r}
pchisq(W_stat, 1, lower.tail = F)


set.seed(414)
values <- c()
pi_null <- (1 - 2*0.3) / (1-0.3)

for(i in 1:10000){
  t_star <- rbinom(1, 170, pi_null)
  psi_star <- (170 - 2*t_star) / (170-t_star)
  wstar <- 2*(log_lik_func(psi_star, 170, t_star) - log_lik_func(0.3,170, t_star))
  values <- c(values, wstar)
  
}


emp_p_value <- mean(values >= W_stat)
emp_p_value
cat("Empirical P value = ", emp_p_value)

ggplot(data = NULL,
mapping = aes(x = values))+
geom_histogram(bins = 40) +
geom_vline(xintercept = W_stat, color = "red") +
labs(title = "Counts of W from Bootstrapping",
     x = "Likelihood Ratio Test Statistic (W) Value", y = "Count")

```


Present your findings.

# Discussion / Conclusion

Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from <https://www.example.com>.

# Appendix

## Code

```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs

If applicable, include detailed mathematical derivations or additional theoretical explanations.
