---
title: "BNT162b2 Decieving Public or Saving Lives?"
author: 'Krish Doshi, Lauren Yee, Kian Sharifzadeh'
date: '`r Sys.Date()`'
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
- \usepackage{colortbl}
- \usepackage{float}
- \usepackage{graphicx}
- \usepackage{booktabs}
- \usepackage{hyperref}
- \usepackage{caption}
- \usepackage{subcaption}
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
#install.packages("janitor")
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
library(janitor)
library(fastR2)

data <- data.frame("Covid_Case" = c(8, 162, 170), "No_Covid_Case" = c(17403, 17349, 34752), "Total"= c(17411, 17511, 34922))

row.names(data) <- c("BNT162b2", "Placebo", "Total")

extended_df <- data.frame("Treatment"=c(rep("BNT162b2", 17411), rep("Placebo", 17511)), "Has_Covid" = c(rep("Yes", 8), rep("No", 17403), rep("Yes", 162), rep("No", 17349)))
```

# Abstract

Write your abstract here.

# Keywords

*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background

Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

## Visualization



# ```{r table}
# table_test <- extended_df %>% 
# tabyl(Group, Has_Covid) %>%
# adorn_totals(where = c("row", "col")) %>%
# adorn_title()
# 
# 
# ```
# 
# ```{r boxplot}
# ggplot(data = extended_df, mapping=aes(x=Has_Covid, fill=Treatment)) +
#   geom_bar(position="fill")
# 
# ```
# 
# ```{r mosaic-plot}
# mosaicplot(table_test, main = "Covid Cases", color = TRUE)
# mosaicplot(~ Group + Has_Covid, data = extended_df, color = TRUE)
# ```



# Statistical Methods



## Model

## Make sure to prove why T is binomial in Appendix (see last slides for example slides)

```{r}
pfizer_estimator <- 8/17411
placebo_estimator <- 162 / 17511

pfizer_estimator
placebo_estimator

relative_risk <- pfizer_estimator / placebo_estimator
relative_risk

psi <- 1 - relative_risk
psi

```

From these calculations, we find that $\pi_{v} = \frac{8}{17411}$ and $\pi_{p} = \frac{162}{17511}$. In other words, $\pi_{v} \approx 0.00046$ and $\pi_{p} \approx 0.00925$.

Pfizer vaccine estimator: 8/17411

Placebo estimator: 162 / 17511

Total: 170 / 34922

Describe the statistical model used.

Let $T \sim Binom(170, \pi)$, where $\pi$ is equivalent to P(Vaccine|Infection). T denotes the number of COVID cases in the vaccine group out of all the sample cases. 170 represents the total number of COVID cases in the sample. Since $n_1$ is approximately equal to $n_2$, we can rewrite P(Vaccine|Infection) as $\pi = \frac{\pi_{v}}{\pi_{v} + \pi_{p}}$. Our parameter of interest is $\psi$, which is equal to $1 -\frac{\pi_v}{\pi_p} = \frac{\pi_p - \pi_v}{\pi_p}$. Using our value of $\pi$, we can rewrite $\psi$ as $\frac{1-2\pi}{1-\pi}$.

Proof: 

$\frac{1-2\pi}{1-\pi} = \frac{1 - 2(\frac{\pi_{v}}{\pi_{v} + \pi_{p}})}{1- \frac{\pi_{v}}{\pi_{v} + \pi_{p}}}$

$\frac{\pi_v + \pi_p - 2\pi_v}{\pi_v + \pi_p - \pi_v}$

$\frac{\pi_p - \pi_v}{\pi_p} = \psi$

Let $t$ be the number of COVID infections from the vaccine group. We observe $t = 8$. We want to test: $H_0: \psi = 0.3$, $H_1: \psi \ne 0.3$.



## Likelihood Inference

Detail the likelihood approach.

First, we need to find our likelihood function $L^{\ast}(\psi)$. Using our parameterization $\frac{1-2\pi}{1-\pi} = \psi$:

$1-2\pi = \psi(1-\pi)$

$1-2\pi = \psi-\psi\pi$

$1-\psi = \pi(2 -\psi)$

$\pi = \frac{1 -\psi}{2 -\psi}$

We know that $L^{\ast}(\pi)  = \binom{n}{t} (\pi)^{t} \times (1-\pi)^{n-t}$. In this case, $n$ represents the number of people included in the sample and $t$ is the number of people who got infected with COVID. Using $\pi$ in terms of $\psi$, we get: $L^{\ast}(\psi) = \binom{n}{t} (\frac{1 -\psi}{2 -\psi})^{t} \times (1-\frac{1 -\psi}{2 -\psi})^{n-t}$

Next, we solve for $\hat{\psi}^{mle}$. In order to do this, we need to find the log-likelihood function of $L^{\ast}(\psi)$: 

$\ell(\psi) = \ln{\left[\binom{n}{t} (\frac{1 -\psi}{2 -\psi})^{t}\right]} + \ln\left[{(1-\frac{1 -\psi}{2 -\psi})^{n-t}}\right]$

$\ell(\psi) = \ln{\binom{n}{t}} + t\ln{(1-\psi)} -t\ln{(2 - \psi)} + (n-t)\ln{(\frac{2-\psi - (1-\psi)}{2-\psi})}$

$\ell(\psi) = \ln{\binom{n}{t}} + t\ln{(1-\psi)} -t\ln{(2 - \psi)}  + (n-t)\ln{1} - (n-t)\ln{(2-\psi)}$

Now, we take the first derivative with respect to $\psi$:

$\frac{d}{d \psi}\ell(\psi) = 0 - \frac{t}{1-\psi} + \frac{t}{2-\psi} + \frac{n-t}{2-\psi} = -\frac{t}{1-\psi} + \frac{t + (n-t)}{2-\psi}$

$\frac{d}{d \psi}\ell(\psi) = -\frac{t}{1-\psi} + \frac{n}{2-\psi}$

Setting $\frac{d}{d \psi}\ell(\psi) = 0$: $-\frac{t}{1-\psi} + \frac{n}{2-\psi} = 0$

$\frac{t}{1-\psi} = \frac{n}{2-\psi}$

$2t - t\psi = n - n\psi$

$n- 2t = \psi(n - t)$

Therefore, $\boxed{\hat{\psi}^{mle} = \frac{n - 2t}{n-t}}$, which can also be rewritten as $\frac{1 - 2\hat{\pi}^{mle}}{1 - \hat{\pi}^{mle}}$.

To see if $\hat{\psi}^{mle}$, we do the 2nd derivative test:

$\frac{d^2}{d \psi^2}\ell(\psi) = \frac{t}{(1-\psi)^2} -\frac{n}{(2-\psi)^2}$

Depending on the values of $n$ and $t$, our estimate will not always be considered a global maximum. In our case, $\frac{d^2}{d \psi^2}\ell(\psi)$ is negative, which means $\hat{\psi}^{mle}$ is a \fbox{Maximum}.

## Bayesian Inference

Detail the Bayesian approach.

# Results

## Frequentist Way

### Confidence Interval

The equation for the large sample confidence interval of $\psi$ is $\hat{\psi}^{mle} \pm z_{\alpha/2} \hat{SE}$, where $\hat{SE} = \frac{1}{\sqrt{I(\hat{\psi}^{mle})}}$

Using Theorem 12.1 under regularity conditions, we know that $I(\hat{\psi}^{mle}) = E\left[-\frac{d^2}{d\psi^2}\ell(\psi)\right]$

$\frac{d^2}{d \psi^2}\ell(\psi) = -\frac{t}{(1-\psi)^2} + \frac{n}{(2-\psi)^2}$

$E\left[-(-\frac{T}{(\psi - 1)^2} + \frac{n}{(\psi - 2)^2} )\right] = \frac{E\left[T\right]}{(1-\psi)^2} - \frac{n}{(2-\psi)^2}$

$E[T] = n\pi = n\frac{1-\psi}{2-\psi}$

$I(\hat{\psi}^{mle}) = n\frac{1-\psi}{2-\psi} \cdot \frac{1}{{(1-\psi)^2}} - \frac{n}{(2-\psi)^2} = \frac{n}{(2-\psi)(1-\psi)}   - \frac{n}{(2-\psi)^2}$

$\frac{n(2-\psi)}{(2-\psi)^2(1-\psi)} - \frac{n(1-\psi)}{(2-\psi)^2(1-\psi)} = \frac{n}{(2-\psi)^2(1-\psi)} = I(\hat{\psi}^{mle})$

Since we got $I(\hat{\psi}^{mle})$, our standard error is $\frac{1}{\sqrt{\frac{n}{(2-\psi)^2(1-\psi)}}} = \sqrt{\frac{(2-\psi)^2(1-\psi)}{n}}$.

Using all the information given, we can solve the 95\% confidence interval with $t = 8$ and $n = 170$: 

```{r}
n <- 170
t <- 8
psi_mle <- (n - 2*t) / (n - t)
psi_se <- sqrt(((1-psi_mle)*(2-psi_mle)^2) / n)

lower <- psi_mle - qnorm(0.975)*psi_se
upper <- psi_mle + qnorm(0.975)*psi_se

lower
upper
psi_se 

```

$\ell(\psi) = \ln{\binom{n}{t}} + t\ln{(1-\psi)} -t\ln{(2 - \psi)}  + (n-t)\ln{(\frac{1}{2-\psi})}$
```{r}
loglik.psi<-function(psi,x, n){
if(psi >= 1) NA
else{  
  log(choose(n, x)) + x*log(1-psi) - x*log(2-psi) +(n-x)*log(1 / (2-psi))
}
}

ml.psi <- maxLik2(loglik=loglik.psi,
start = 0.3, x = 8,n = 170)

plot(ml.psi) +
labs(title = "Second order approximation for Psi (Parametrization for Pi)",
subtitle = "n = 170, t = 8",
x = expression(psi))



```

```{r}
set.seed(414)

pi_samples <- rbinom(n = 100000, size = 170, prob =  8/170)
pi_mle <- pi_samples / 170
psi_samples <- (1 - 2*pi_mle) / (1-pi_mle)
sd(psi_samples)

```

### Likelihood Ratio Test Statistic

The likelihood ratio statistic is calculated by $W = 2\ln\Lambda$, which is equivalent to $2\ln(\frac{L(\hat{\psi}^{mle})}{L(\hat{\psi}^{null})}) = 2\left[\ln{L(\hat{\psi}^{mle})} - \ln{L(\hat{\psi}^{null})}\right]$. Given that $\hat{\psi}^{mle} \approx 0.9506$ and $H_0: \psi = 0.3$, we are solving for $2\left[\ell{(0.9506)} - \ell{(0.3)}\right]$. Using that $t = 8$ and $n = 160$, we can calculation $\ell(0.9506)$ and $\ell(0.3)$:

$\ell(0.9506) = \ln{\binom{160}{8}} + 8\ln{(1-0.9506)} -8\ln{(2 - 0.9506)}  + (162)\ln{(\frac{1}{2-0.9506})} \approx -1.945$

$\ell(0.3) = \ln{\binom{160}{8}} + 8\ln{(0.7)} -8\ln{(1.3)}  + (162)\ln{(\frac{1}{1.3})} \approx - 62.746$

Then, $W = 2(-1.945 + 62.746) \approx \boxed{121.601}$


Likelihood Ratio Test Statistic
```{r}
log_lik_func <- function(psi, n , t){
  log(choose(n, t)) + t*log(1-psi) - t*log(2-psi) -(n-t)*log((2-psi))
}

log_mle <- log_lik_func(psi_mle, 170, 8)
log_null <- log_lik_func(0.3, 170, 8)

log_mle
log_null

W_stat <- 2*(log_mle - log_null)
W_stat

```

```{r}
pchisq(W_stat, 1, lower.tail = F)


set.seed(414)
values <- c()

for(i in 1:10000){
  t_star <- rbinom(30, 170, 0.3)
  pi_mle <- t_star / 170
  psi_star <- (170 - 2*t_star) / (170-t_star)
  wstar <- 2*(log_lik_func(psi_star, 170, t_star) - log_lik_func(0.3,170, t_star))
  values <- c(values, wstar)
  
}
# sim_func <- function(B){
#   t_star <- rbinom(B, 170, 0.3)
#   pi_mle <- t_star / 170
#   psi_star <- (1 - 2*pi_mle) / (1-pi_mle)
#   wstar <- 2*(log_lik_func(psi_star, 170, t_star) - log_lik_func(0.3,170, t_star))
#   wstar
#   return(wstar)
# }
# 
# sim_func(B)

# null_sim <- lapply(1:B, sim_func)
# Wstar <- c(unlist(null_sim))
# which.min(Wstar)

emp_p_value <- mean(values >= W_stat)
emp_p_value
# cat("Empirical P value = ", emp_p_value)

# ggplot(data = NULL,
# mapping = aes(x = Wstar))+
# geom_histogram() +
# geom_vline(xintercept = W_stat) +
# labs(title = "Counts of W from Bootstrapping", 
#      x = "Likelihood Ratio Test Statistic (W) Value", y = "Count")

```


Present your findings.

# Discussion / Conclusion

Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from <https://www.example.com>.

# Appendix

## Code

```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs

If applicable, include detailed mathematical derivations or additional theoretical explanations.
